---
title: "Practical Machine Learning Course Project Report"
author: "Anita"
date: "December 24, 2017"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
cache= TRUE
```

###Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Data Sources
The training data for this project is available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data is available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
The data for this project comes from this original source: http://groupware.les.inf.puc-rio.br/har.

###Objective
The objective of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. 
Any of the other variables can be used to predict with. This report describes how the model was built, how cross validation was used, what is the expected out of sample error is and why certain choices were made. This prediction model will be used to predict 20 different test cases.

###Reproducibility
In order to reproduce the same results, you need to install/load the set of packages loaded below and set seed equal to the one I have used.

```{r loadlib}
library(rattle)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)
library(randomForest)
library(RColorBrewer)
set.seed(56789)

     cache = TRUE
```

### Data Loading
In this setp we will download the data files from the URL and load them into testing and training Data Frames.

```{r loaddata}
setwd("C://Files/Data Science/Machine Learning/Project")
trainUrl  <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl   <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "training.csv"
testFile  <- "testing.csv"
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile = trainFile)
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile = testFile)
}
#Load the data into data frames
rm(trainUrl)
rm(testUrl)
trainDF <- read.csv(trainFile)
testDF <- read.csv(testFile)
dim(trainDF)
dim(testDF)

```

###Data Cleanup

In this step, we will clean the dataset and get rid of observations with NA values,redundant variables and  near zero variance values.
```{r removeredundant}
#Removing NA values
condition <- (colSums(is.na(trainDF)) == 0)
training  <- trainDF[, condition]
testing   <- testDF[, condition]
rm(condition)
rm(trainDF)
rm(testDF)

#Removing redundant variables
regex    <- grepl("^X|timestamp|user_name", names(training))
training <- training[, !regex]
testing  <- testing[, !regex]
rm(regex)
dim(training)
dim(testing)

#Removing near zero variances
NZV <- nearZeroVar(training, saveMetrics = TRUE)
head(NZV, 20)
training <- training[, !NZV$nzv]
testing  <- testing[, !NZV$nzv]
dim(training)
dim(testing)

rm(NZV)


#Correlation Matrix of Columns in the Training Data set.

corrplot(cor(training[, -length(names(training))]), method = "color", tl.cex = 0.5)
```

###Partitioning the training data set 
In this step we will partition the training data set into training(70%) and validation(30%) data sets

```{r partition}
set.seed(56789) #For reproducibility
inTrain    <- createDataPartition(training$classe, p = 0.70, list = FALSE)
training   <- training[inTrain, ]
validation <- training[-inTrain,]


#The Dataset now consists of:
#No of observations in Training Data: 
dim(training)
#2. No of observations in Validation Data:
dim(validation)
#3. No of observations in Testing Data:
dim(testing)
```

###Data Modelling

#### Decision Tree
Fit a predictive model for activity recognition using Decision Tree algorithm.

```{r decisiontree}
modelDT <- rpart(classe ~ ., data = training, method = "class")
prp(modelDT)

#Estimate the performance of the model on the validation data set
predictDT <- predict(modelDT, validation, type = "class")
confusionMatrix(validation$classe, predictDT)
accuracy <- postResample(predictDT, validation$classe)
ose <- 1 - as.numeric(confusionMatrix(validation$classe, predictDT)$overall[1])
rm(predictDT)
rm(modelDT)
```

#### Random Forest
Fit a predictive model for activity recognition using Random Forest algorithm(using 5-fold cross validation).

```{r randomforest}

modelRF <- train(classe ~ ., data = training, method = "rf", trControl = trainControl(method = "cv", 5), ntree = 250)
modelRF
#Estimate the performance of the model on the validation data set
predictRF <- predict(modelRF, validation)
confusionMatrix(validation$classe, predictRF)
accuracy <- postResample(predictRF, validation$classe)
ose <- 1 - as.numeric(confusionMatrix(validation$classe, predictRF)$overall[1])
rm(predictRF)

#The Estimated Accuracy of the Random Forest Model is:
accuracy 
#The Estimated Out-of-Sample Error is:
ose

#Random Forests yielded better results.
```

###Predicting with testing data set
```{r predict}
rm(accuracy)
rm(ose)
predict(modelRF, testing[, -length(names(testing))])
```


